### 2023-10-17 doc/Thesis/chapter/031-theory-calculations.tex
corrected until comp details

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine Learning and Statistics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Why introduce machine learning to this project? 
%I wanted to apply and delve into what I've been studying during my undergrad courses. 
%This seemed like the perfect oppurtunity. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Is machine learning just hyped statistics? No! 
		%{https://doi.org/10.1177\%2F1352458520978648}
		%{Machine and deep learning in MS research are just powerful statistics – No}
%What is \gls{ml}? 
Some might argue that \gls{ml} is just overglorified statistics, but that's not the case. 
Although \gls{ml} uses several statistical methods, their goal and frame conditions are different. 
%\td{The goal of machine learning is to ???}
%
\Gls{ml} tries to predict unseen data points accurately and statistics is a subfield of maths 
which tries to get insight into a given data. 
For example, in a statistical model, it is desirable to reduce the number of inputs. 
This allows a statistician to better study how a change in input variables can 
directly affect a output variable\cite{gontcharov2019}.
%pic from \url{https://towardsdatascience.com/notes-on-artificial-intelligence-ai-machine-learning-ml-and-deep-learning-dl-for-56e51a2071c2}
More precisely, mathematical statistics (stochastic)\cite{haertler2014statistisch} is the subject of 
finding mathematical models to describe the data 
whereas (classical) statistics is the domain of representing the data. 
Nonetheless, statistics and maths is at the very basis of \gls{ml}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{EMMA}
\Gls{emma} is an implementation of \gls{pso} in the \texttt{R} programming language. 
Each time step the dependent variables for all possible input variable combinations %$\mathbf{x} = \{x_1,\cdots ,x_N\}$ 
are predicted with the help of \gls{mars}. 
The \gls{mars} \gls{rf} is then used to chose the next position for each particle.
\Gls{mars} is a regression method introduced by Friedman's 1991 paper 
"Multivariate Adaptive Regression Splines"\cite{friedman1991multivariate}. 
The \texttt{CRAN} package which implements \gls{mars} is called \texttt{earth} due to \gls{mars} being trade marked\cite{mars}.
%%% MARS 
\gls{mars} is an extension of multivariate linear regression.
Friedman presented \gls{mars} as an alternative to piecewise polynomials (splines) and local averaging methods (e.g. kernel functions), whose number of parameter quickly exceeds the practical for moderate dimensional problems ($n>5$). 
%
%to higher dimensions (n > 2) is straightforward in principle but difficult in practice. 
%"These difficulties are related to the so-called “curse-of-dimensionality”, a phrase coined by Bellman (1961) to express the fact that exponentially increasing numbers of points are needed to densely populate Euclidean spaces of increasing dimension."\cite{friedman1988fitting}
%
The main advancements of \gls{mars} is the ability to fit (possibly complex) interactions and non-linearities without losing interpretability. 
%
%- "The key ingredient that advances this approach to general settings is the ability to fit (possibly complex) interactions among the variables through the product terms that are permitted to enter the approximation (9), if required by the fit."\cite{friedman1988fitting}
%% MARS MODEL SELECTION
%- "The approximation is developed in a forward/backwards stepwise recursive manner in analogy with the recursive partitioning approach. "\cite{friedman1988fitting}
%
%%% FORWARD MARS
The \gls{rf} is developed in a forward/backward stepwise recursive manner\cite{friedman1988fitting}. 
During the forward recursion terms are added in pairs until a certain number of terms is reached. 
Each pair consists of two hinge functions (also called rectifier functions) multiplied with a term already part of the \gls{rf} (including the constant term).
The maximum degree of interaction is 2 for \gls{emma}, but can be varied. 
This means that only two basis functions can be multiplied (excluding the constant term) to form a subsequent term.
Pairwise added hinge functions are of the simple form $h(x-c)$ and $h(c-x)$ where $h(e)= max(0,e)$ and with $x$ being an independent variable, $c$ being a constant and $e$ being any expression. 
%%% BACKWARD EMMA
Then, the backwards algorithm regularizes the function by removing individual terms. 
The metric which decides if a term should be removed is called \gls{gcv} and was introduced by Wahba and Cravenin 1969\cite{wahba1979smoothing}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine Learning and Statistics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Why introduce machine learning to this project? 
%I wanted to apply and delve into what I've been studying during my undergrad courses. 
%This seemed like the perfect oppurtunity. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Is machine learning just hyped statistics? No! 
		%{https://doi.org/10.1177\%2F1352458520978648}
		%{Machine and deep learning in MS research are just powerful statistics – No}
%What is \gls{ml}? 
Some might argue that \gls{ml} is just overglorified statistics, but that is not the case. 
Although \gls{ml} uses several statistical methods, their goal and frame conditions are different. 
%\td{The goal of machine learning is to ???}
%
\Gls{ml} tries to predict unseen data points accurately and statistics is a subfield of maths 
which tries to get insight into a given data. 
For example, in a statistical model, it is desirable to reduce the number of inputs. 
This allows a statistician to better study how a change in input variables can 
directly affect a output variable\cite{gontcharov2019}.
%pic from \url{https://towardsdatascience.com/notes-on-artificial-intelligence-ai-machine-learning-ml-and-deep-learning-dl-for-56e51a2071c2}
More precisely, mathematical statistics (stochastic)\cite{haertler2014statistisch} is the subject of 
finding mathematical models to describe the data 
whereas (classical) statistics is the domain of representing the data. 
Nonetheless, statistics and maths is at the very basis of \gls{ml}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{EMMA}
\Gls{emma} is an implementation of \gls{pso} in the \texttt{R} programming language. 
Each time step the dependent variables for all possible input variable combinations %$\mathbf{x} = \{x_1,\cdots ,x_N\}$ 
are predicted with the help of \gls{mars}. 
The \gls{mars} \gls{rf} is then used to chose the next position for each particle.
\Gls{mars} is a regression method introduced by Friedman's 1991 paper 
"Multivariate Adaptive Regression Splines"\cite{friedman1991multivariate}. 
The \texttt{CRAN} package which implements \gls{mars} is called \texttt{earth} due to \gls{mars} being trade marked\cite{mars}.
%%% MARS 
\gls{mars} is an extension of multivariate linear regression.
Friedman presented \gls{mars} as an alternative to piecewise polynomials (splines) and local averaging methods (e.g. kernel functions), whose number of parameter quickly exceeds the practical for moderate dimensional problems ($n>5$). 
%
%to higher dimensions (n > 2) is straightforward in principle but difficult in practice. 
%"These difficulties are related to the so-called “curse-of-dimensionality”, a phrase coined by Bellman (1961) to express the fact that exponentially increasing numbers of points are needed to densely populate Euclidean spaces of increasing dimension."\cite{friedman1988fitting}
%
The main advancements of \gls{mars} is the ability to fit (possibly complex) interactions and non-linearities without losing interpretability. 
%
%- "The key ingredient that advances this approach to general settings is the ability to fit (possibly complex) interactions among the variables through the product terms that are permitted to enter the approximation (9), if required by the fit."\cite{friedman1988fitting}
%% MARS MODEL SELECTION
%- "The approximation is developed in a forward/backwards stepwise recursive manner in analogy with the recursive partitioning approach. "\cite{friedman1988fitting}
%
%%% FORWARD MARS
The \gls{rf} is developed in a forward/backward stepwise recursive manner\cite{friedman1988fitting}. 
During the forward recursion, terms are added in pairs until a certain number of terms is reached. 
Each pair consists of two hinge functions (also called rectifier functions) multiplied with a term already part of the \gls{rf} (including the constant term).
The maximum degree of interaction is 2 for \gls{emma}, but can be varied. 
This means that only two basis functions can be multiplied (excluding the constant term) to form a subsequent term.
Pairwise added hinge functions are of the simple form $h(x-c)$ and $h(c-x)$ where $h(e)= max(0,e)$ and with $x$ being an independent variable, $c$ being a constant and $e$ being any expression. 
%%% BACKWARD EMMA
Then, the backwards algorithm regularizes the function by removing individual terms. 
The metric which decides if a term should be removed is called \gls{gcv} and was introduced by Wahba and Cravenin 1969\cite{wahba1979smoothing}.

### 2023-10-17 doc/Thesis/chapter/033-linreg-krr-svm.tex
corrected until comp details

%\textbf{Kernel Ridge Regression}
%%%%%%%%%%%%%%%%%%%%%%%%%%% { Kernel Ridge Regression } KRR %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{\Gls{krr}} combines ridge regression with the kernel method. 
A~kernel transforms data in such a way that a linear hyperplane (a point in one dimension, a line in two dimensions, 
a plane in three dimensions) can fit data in regression problems or separate the data in classification problems without actually doing the transormation for every data point, which lowers computational costs.
A kernel is some kind of similarity measure 
which fulfills the requirements of nonnegativity, symmetry and linearity\cite{rupp2015machine}.
%\enlargethispage{\baselineskip}
%\looseness=-1
The following equations show definitions for 
linear~(eq.~\ref{eq:lin-kernel}), 
polynomial (eq.~\ref{eq:pol-kernel}), 
sigmoidal (eq.~\ref{eq:sig-kernel}) and 
radial basis functionial kernel~(eq.~\ref{eq:rbf-kernel}), with $\gamma$ as fixed hyperparameter and $c_0$ as parameter to optimize.
A requirement for using kernels is having a dot product in the loss function. 
This can be accomplished by expressing $\mathbf{k}$ in terms of $\mathbf{X}$: $\mathbf{k}=\mathbf{X}^\top \mathbf{r}$
\cite{rudin2020least}.

@@

The \gls{svm} was initially developed by Vladimir Vapnik for the binary classification of separable data, then improved to handle non-separable data and eventually adapted to solve regression problems.

@@

Thus maximizing the margin is equivalent with minimizing $\|\mathbf{w}\|$ and minimizing $\frac{1}{2}\|\mathbf{w}\|^2$ (mathematical convenience for further steps).

@@

mathemagically arrive at the loss function which should be maximized: 

@@

to zero and inserting into eq. (\ref{eq:svm03}))\cite{winston1992artificial,cherkassky1998learning}: 

@@

	L_{SM} = \frac{1}{2} \|\mathbf{w}\|^2 + \sum_i^n max(0, 1- y_i ( \mathbf{w} \mathbf{x}_i + b ) 

@@

If the prediction $\mathbf{w}\mathbf{x}_i+b$ and the true category $y_i$ don't agree, 

@@

The substraction of a negative number will result in a positive penalizization. 

@@

the substraction will lead to a negative number and the maximum will be 0, if the sample is ousite of the margin. 

@@

but also by complex decision boundaries due to the kernel trick. 

>>>>

%\textbf{Kernel Ridge Regression}
%%%%%%%%%%%%%%%%%%%%%%%%%%% { Kernel Ridge Regression } KRR %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{\Gls{krr}} combines ridge regression with the kernel method. 
A~kernel transforms data in such a way that a linear hyperplane (a point in one dimension, a line in two dimensions, 
a plane in three dimensions) can fit data in regression problems or separate the data in classification problems without actually doing the transformation for every data point, which lowers computational costs.
A kernel is some kind of similarity measure 
which fulfills the requirements of nonnegativity, symmetry and linearity\cite{rupp2015machine}.
%\enlargethispage{\baselineskip}
%\looseness=-1
The following equations show definitions for 
linear~(eq.~\ref{eq:lin-kernel}), 
polynomial (eq.~\ref{eq:pol-kernel}), 
sigmoidal (eq.~\ref{eq:sig-kernel}) and 
radial basis functionial kernel~(eq.~\ref{eq:rbf-kernel}), with $\gamma$ as fixed hyperparameter and $c_0$ as parameter to optimize.
A requirement for using kernels is having a dot product in the loss function. 
This can be accomplished by expressing $\mathbf{k}$ in terms of $\mathbf{X}$: $\mathbf{k}=\mathbf{X}^\top \mathbf{r}$
\cite{rudin2020least}.

@@

The \gls{svm} was initially developed by Vladimir Vapnik for the binary classification of separable data, then improved to handle non-separable data\cite{cortes1995support} and eventually adapted to solve regression problems\cite{drucker1996support}.

@@

Thus, maximizing the margin is equivalent with minimizing $\|\mathbf{w}\|$ and minimizing $\frac{1}{2}\|\mathbf{w}\|^2$ (mathematical convenience for further steps).

@@

\textit{mathemagically} arrive at the loss function which should be maximized: 

@@

to zero and inserting into eq. (\ref{eq:svm03}))\cite{winston1992artificial,cherkassky1998learning} in the following way: 

@@

	L_{SM} = \frac{1}{2} \|\mathbf{w}\|^2 + \sum_i^n max \left( 0, 1- y_i ( \mathbf{w} \mathbf{x}_i + b ) \right) 

@@

If the prediction $(\mathbf{w}\mathbf{x}_i+b)$ and the true category $y_i$ don't agree, 

@@

The substraction of a negative number will result in a positive penalization. 

@@

the substraction will lead to a negative number and the maximum will be 0 if the sample is ousite of the margin. 

@@

but also by intricate decision boundaries due to the kernel trick. 
